---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%"
)
```

# Sad music and hormones

## Data and scripts 

The following are all statistical analyses reported in the paper. All data files are available at Github repository: [https://github.com/tuomaseerola/sad-music-hormones](https://github.com/tuomaseerola/sad-music-hormones)

_Created: 9/4/2021_.


## Data files
* `background.csv`: Individual differences variable (`QoL`, `TAS`, `MSI`, etc.) with unique `ID` for each participant.
* `emotions.csv`: Self-reported emotions (`moved`, `relaxed`, `sad`, `anxious`, `pleasantness`).
* `mood_and_emotions.csv`: Mood and emotions (`pos` and `neg` moods).
* `aggregate_data.csv`: Hormone variables (`PRL`, `OXY`, `CRT`, `ACTH`, `CRP`) and participants IDs and Conditions.

* `EDA.csv`: Electrodermal variables (`CDA.ISCR`,`CDA.nSCR`,`CDA.AmpSum`,`CDA.Tonic`) and participants IDs and Conditions.
* `HRV.csv`: Heart rate variability predictors (`SDNN`,`LF`,`HF`,`LF/HF`) and participants IDs and Conditions.

## R Scripts

* `analysis_moods.R`: Reads and preprocesses data (relative change, outliers), and runs GLMM and produces planned contrasts.
* `analysis_emotions.R`: Reads and preprocesses data (relative change, outliers), and runs GLMM and produces planned contrasts. 
* `analysis_EDA.R`: Reads and preprocesses data (relative change, outliers), and runs GLMM and produces planned contrasts.
* `analysis_HRV.R`: Reads and preprocesses data (relative change, outliers), and runs GLMM and produces planned contrasts.
* `figure1.R`: Combines the plots created by the other scripts and produces a figure.

# Analysis of moods

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}
source('analysis_moods.R')
```

# Analysis of emotions

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}
source('analysis_emotions.R')
```

# Analysis of hormones

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}
source('analysis_hormones.R')
```

## Summary of hormonal variables

A full breakdown of all variables is given in Supporting Information Table S1.

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}

library(dplyr)
m <- df_rel %>%
  dplyr::group_by(EMPATHY,Condition) %>%
  dplyr::summarise(n=n(),PRL_M=mean(PRL,na.rm = TRUE),PRL_SE=sd(PRL,na.rm = TRUE)/sqrt(n), OXY_M=mean(OXY,na.rm = TRUE),OXY_SE=sd(OXY,na.rm = TRUE)/sqrt(n), CRT_M=mean(CRT,na.rm = TRUE),CRT_SE=sd(CRT,na.rm = TRUE)/sqrt(n), ACTH_M=mean(ACTH,na.rm = TRUE),ACTH_SE=sd(ACTH,na.rm = TRUE)/sqrt(n), CRP_M=mean(CRP,na.rm = TRUE),CRP_SE=sd(CRP,na.rm = TRUE)/sqrt(n))

knitr::kable(m,digits = 3,caption = 'Table S1. Means and standard errors of all hormone markers.')

```

## Additional factors

Two additional covariates were added to the main analysis, hormonal contraception and menstruation cycle. Hormonal contraception was used by 28 participants and it is coded as a binary variable. Information about the phase of the participant's menses is classified from the days since menses into five categories based on prior research [@andreano2008menstrual]. The table also shows you how many of our participants fall in each phase.

Table S2. Summary of the phase of the mense in participants.

| Phase | Definition | N |
|------:|-----------:|--:|
| early-follicular phase | days 1–7 since menstruation| 6 |
| late-follicular | days 8–17 since menstruation| 21 |
| mid-luteal |days 18–24 since menstruation| 16 |
| late-luteal |days 25-29 since menstruation| 16 |
| NA |not applicable, not known| 3 |

Both of these variables have been added as random factors in the analysis of prolactinand oxytocin below. As the results without these variables is retained in the comments, it is easy to see that these variables do materially change the outcome of the analyses (there are marginal changes in the p values).

### Impact on prolactin

```{r,echo=TRUE,warning=FALSE,message=FALSE,include=TRUE, cache=TRUE}
m1_additional <- lme4::lmer(PRL ~ Condition * EMPATHY + (1|Hormone) + (1|MenstPhase) + (1|ID) + (1|Session), data=df_rel) # relative
em2 <- emmeans::emmeans(m1_additional,specs = ~ Condition * EMPATHY) #

# This is Between Group comparison                                     estimate     SE df t.ratio p.value
knitr::kable(emmeans::contrast(em2,method=list(Silence_High - Silence_Low)))#    -2.39 1.73 71.8 -1.381  0.1716 
knitr::kable(emmeans::contrast(em2,method=list(Music_High - Music_Low)))    #    -4.25 1.73 71.8 -2.454  0.0166  
# Within group comparison 
knitr::kable(emmeans::contrast(em2,method=list(Silence_High - Music_High))) #       2.3 0.727 59 3.168   0.0024 
knitr::kable(emmeans::contrast(em2,method=list(Silence_Low - Music_Low)))   #     0.445 0.751 59 0.592   0.5562
```

### Impact on oxytocin
```{r,echo=TRUE,warning=FALSE,message=FALSE,include=TRUE, cache=TRUE}
m1_additional <- lme4::lmer(OXY ~ Condition * EMPATHY + (1|Hormone) + (1|MenstPhase) + (1|ID) + (1|Session), data=df_rel) # relative
em2 <- emmeans::emmeans(m1_additional,specs = ~ Condition * EMPATHY) #
# This is Between Group comparison                                     estimate     SE df t.ratio p.value
knitr::kable(emmeans::contrast(em2,method=list(Silence_High - Silence_Low)))#   3.16 4.4 104 0.717   0.4750 
knitr::kable(emmeans::contrast(em2,method=list(Music_High - Music_Low)))    #  -9.7 4.4 104 -2.205  0.0297
# WITHIN GROUPS
knitr::kable(emmeans::contrast(em2,method=list(Silence_High - Music_High))) #  8.03 3.39 59.1 2.370   0.0211 
knitr::kable(emmeans::contrast(em2,method=list(Silence_Low - Music_Low)))   # -4.83 3.5 59.1 -1.381  0.1724
```

# Analysis of psychophysiology

## Analysis of EDA

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}
source('analysis_EDA.R')
```

## Analysis of HRV

```{r,echo=TRUE,results='asis',warning=FALSE,message=FALSE}
source('analysis_HRV.R')
```

# Create Figure 1

```{r, echo=T,warning=FALSE,message=FALSE,fig.width=12,fig.height=5}
source('Figure1.R')
```

# References

<div id="refs"></div>


# Appendix

```{r, echo=T}
sessionInfo()
```

